---
title: "Project 2"
output: pdf_document
---


## 1: Synopsis
The goal of the assignment is to explore the NOAA Storm Database and 
explore the effects of severe weather events on both population and economy.
The time period for the database is between 1950 and November 2011.

Following analysis investigates which types of severe weather events are most harmful on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

## 2: Data Processing

### 2.1: Data Loading

Download the raw data file and extract the data into a dataframe.Then convert to a data.table
```{r DataLoading, echo = F}
# Load libraries
library(tidyverse)
library(ggplot2)
library(reshape2)
# Load data
storm <- read.csv('repdata_data_StormData.csv')
```

### 2.2: Examining Column Names

```{r ColumnNames}
colnames(storm)
```


### 2.3: Data Subsetting

Subset the dataset on the parameters of interest. Basically, we remove the columns we don't need for clarity. 
```{r DataSubsetting, results="hide"}
# Only use data where fatalities or injuries occurred.  
storm <- storm %>%
    filter(EVTYPE != "?" & 
               (INJURIES > 0 | FATALITIES > 0 | 
                    PROPDMG > 0 | CROPDMG > 0)) %>% 
    select( c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG",
              "PROPDMGEXP","CROPDMG", "CROPDMGEXP"))
```

### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)

Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.
```{r CorrectingExponents, results="hide"}
# Map property damage alphanumeric exponents to numeric values.
storm$PROPDMGEXP <-plyr::mapvalues(storm$PROPDMGEXP, 
                              from = c("K", "M","", "B", "m", "+", "0", "5",
                                       "6", "?", "4", "2", "3", "h", "7",
                                       "H", "-", "1", "8"), 
                              to = c(10^3, 10^6, 1, 10^9, 10^6, 0,1,10^5, 10^6,
                                     0, 10^4, 10^2, 10^3, 10^2, 10^7, 10^2,
                                     0, 10, 10^8))

storm$PROPDMGEXP <- as.numeric(as.character(storm$PROPDMGEXP))
storm$PROPDMGTOTAL <- (storm$PROPDMG * storm$PROPDMGEXP)/1000000000

# Map crop damage alphanumeric exponents to numeric values
storm$CROPDMGEXP <- plyr::mapvalues(storm$CROPDMGEXP, 
                              from = c("","M", "K", "m", "B",
                                       "?", "0", "k","2"), 
                              to = c(1,10^6, 10^3, 10^6, 10^9, 
                                     0, 1, 10^3, 10^2))

storm$CROPDMGEXP <- as.numeric(as.character(storm$CROPDMGEXP))
storm$CROPDMGTOTAL <- (storm$CROPDMG * storm$CROPDMGEXP)/1000000000
```


### 2.5: Making Economic Cost Columns

```{r EconomicCostColumns}
storm <- storm %>%
    mutate(propCost = PROPDMG * PROPDMGEXP, 
           cropCost = CROPDMG * CROPDMGEXP)
```

### 2.6: Calcuating Total Property and Crop Cost
```{r}
totalCost <- storm %>%
    group_by(EVTYPE) %>%
    summarize(propCost = sum(propCost), 
              cropCost = sum(cropCost))
head(totalCost)

```
```{r TotalPropertyCropCost}
totalCost <- storm %>%
    group_by(EVTYPE) %>%
    summarize(propCost = sum(propCost), 
           cropCost = sum(cropCost)) %>%
    mutate(Total_Cost = propCost + cropCost) %>%
    select(EVTYPE, Total_Cost) %>%
    arrange(-Total_Cost)

totalCost <- totalCost[1:10, ]
head(totalCost, 5)
```

### 2.7: Calcuating Total Fatalities and Injuries

```{r TotalFatalitiesInjuriesCalc}
totalInjuries <- storm %>%
    group_by(EVTYPE) %>%
    summarize( FATALITIES = sum(FATALITIES), 
            INJURIES = sum(INJURIES)) %>%
    mutate(total = FATALITIES + INJURIES) %>%
    select(EVTYPE, total ) %>%
    arrange(-total)

totalInjuries <- totalInjuries[1:10, ]
head(totalInjuries, 5)
```

## 3: Results

### 3.1: Events that are Most Harmful to Population Health

Melting data.table so that it is easier to put in bar graph format 
```{r HealthResults}
bad_stuff <- melt(totalInjuries, 
                  id.vars="EVTYPE", 
                  variable.name = "bad")
head(bad_stuff, 5)
```
```{r healthChart}
# Create chart
healthChart <- ggplot(bad_stuff, 
                      aes(reorder(EVTYPE, -value), value)) +
    geom_bar(stat="identity", 
             aes(fill=bad), 
             position="dodge")  + 
    ylab("Frequency Count") + 
    xlab("Event Type") + 
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Top 10 US Killers") + 
    theme(plot.title = element_text(hjust = 0.5))

healthChart
```

### 3.2: Events that have the Greatest Economic Consequences

Melting data.table so that it is easier to put in bar graph format 
```{r EconConsequences}
econ_consequences <- melt(totalCost, 
                          id.vars="EVTYPE", 
                          variable.name = "Damage_Type")
head(econ_consequences, 5)
```

```{r econChart}
# Create chart
econChart <- ggplot(econ_consequences, 
                    aes(reorder(EVTYPE, -value), 
                        value)) + 
    geom_bar(stat="identity",
             aes(fill=Damage_Type), 
             position="dodge") + 
    ylab("Cost (dollars)") + 
    xlab("Event Type") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Top 10 US Storm Events causing Economic Consequences") + 
    theme(plot.title = element_text(hjust = 0.5))

econChart
```